Efficient High-Level Abstractions for Web Programming

### 2013-09-02 papers/gpce2013/gpce2013.tex
typos

Writing large Web applications is known to be difficult. One challenge comes from the fact that the
application's logic is scattered into heterogeneous clients and servers, making it difficult to
share code between both sides or to move code from one side to the other. Another challenge is
performance: while Web applications rely on ever more code on the client-side, they may run on smart
phones with limited hardware capabilities. These two challenges raise the following problem: how to
benefit from high-level languages and libraries making code complexity easier to manage and
abstracting over the clients and servers differences without trading this engineering comfort for
performance? This article presents high-level abstractions defined as deep embedded DSLs in Scala,
that can (1) generate efficient code leveraging the target platform characteristics, (2) be shared
between clients and servers. Though code written with our DSL has a high level of abstraction our
benchmark on a real world application reports that it runs as fast as hand tuned low-level
JavaScript code.
\end{abstract}

@@

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that
generates JavaScript code, making it possible to write the client-side code of Web applications
using JavaScript~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the
engineering dilemma described above. This paper enriches js-scala with the following
contributions\footnote{The code is available at
\href{http://github.com/js-scala}{http://github.com/js-scala}}:

@@

We validate our approach with a case study implemented with various candidate technologies and
discuss the relative pro and cons of them. Though the code written in our DSL is high-level and can
be shared between clients and servers, it has the same runtime performances on client-side as
hand-tuned low-level JavaScript code.

@@

Another approach consists in defining several independent domain-specific
languages~\cite{Van00_DSL}, each one focusing on concerns specific to a given problem domain, and
then to combine all the source artifacts written with these language into one executable program, as
shown in figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
the previous approach because each language has a limited set of features. On the other hand, it is
difficult to have interoperability between DSLs. \cite{Groenewegen08_WebDSL} gave an example of such
a domain-specific language for defining Web applications.

@@

A high-level abstraction for searching elements in a document could be just one function
finding all elements matching a given CSS selector. In fact, most JavaScript
developers\footnote{According to
\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
code is both shorter and simpler, thanks to its higher-level of abstraction. jQuery provides an API
that is simpler to master because it has less functions, but this benefit comes at the price of a
decrease in runtime performances.

Instead, we propose a solution that has a high-level API but generates JavaScript code using the
specialized native API, when possible, in order to get both engineering comfort and performances.
We achieve this by analyzing, during the first evaluation step, the selector that is
passed as parameter and, when appropriate, by producing JavaScript code using the specialized API,
and otherwise producing code using \code{querySelector} and \code{querySelectorAll}.

@@

Our goal is to evaluate the level of abstraction provided by our solution and its performances, by
comparing it with common approaches. We take the number of lines of code as an inverse approximation
of the level of abstraction. We also evaluate the ability to share code between client and server
sides.

@@

The runtime performances of the vanilla JavaScript, HaXe and js-scala versions are similar (though
the js-scala version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and
the HaXe versions use low-level code compared to js-scala, as shown in the middle of the figure
(lines of code): the js-scala version needs only 74 lines of code while the vanilla JavaScript
version needs 116 lines of code (57\% bigger) and the HaXe version needs 148 lines of code (100\%
bigger). The jQuery JavaScript version, which code is high-level (54 lines of code, 27\% less
than js-scala) runs 10 times slower than the js-scala version.

@@

We were able to share some DOM fragment definitions between server-side and client-side only in
the js-scala version. In the GWT version we don’t have choice: dynamic DOM fragments are always
built only on client-side (a practice that makes it more difficult to make the pages content
crawlable by search engines and may increase the initial display time). In the other versions the
code for building the DOM fragment is duplicated between client and server sides, representing 20
lines of JavaScript code (17\% of the total) and 15 lines of HTML (5\% of the total) in the
JavaScript version, and 19 lines of HaXe code (13\% of the total) and 15 lines of HTML (5\%
of the total) in the HaXe version. In the js-scala version the DOM fragment definitions shared
between clients and servers represent 22 lines of Scala code (30\% of the total) and save 15 lines
of HTML (5\% of the total).

@@

Our goal was to put the runtime performances in perspective with the level of abstraction. We are
aware that the indicator we chose as an inverse approximation of the abstraction level, the number
of lines of code, is not scientifically established and may be subject to discussion. However, we
think it is a reasonable approximation in our case because all the candidate languages we use have
a similar syntax, inherited from the C programming language.

@@

We implemented our solution as compiled embedded DSLs in Scala. Generating code from our DSLs is a
two steps process: an initial Scala program first evaluates to an intermediate representation of
the final program that is traversed by code generators to produce the final JavaScript code.
Domain-specific optimizations can happen during the IR construction (as shown in section
\ref{sec:selectors}) or during the code generation (as shown in section \ref{sec:option}).

@@

These characteristics allowed us to capture some Web programming patterns as high-level
abstractions, making the code of our application simpler to reason about and making some parts of
the code reusable between client and server sides, while keeping execution performances on
client-side as high as if we used hand-tuned low-level JavaScript code.

@@

Our two kinds of benchmarks, (i) micro-benchmarks to evaluate one abstraction and (ii) a benchmark
on a real application using these abstractions, show performances similar to hand-optimized
low-level code.

>>>>

Writing large Web applications is known to be difficult. One challenge comes from the fact that the
application's logic is scattered into heterogeneous clients and servers, making it difficult to
share code between both sides or to move code from one side to the other. Another challenge is
performance: while Web applications rely on ever more code on the client-side, they may run on smart
phones with limited hardware capabilities. These two challenges raise the following problem: how to
benefit from high-level languages and libraries making code complexity easier to manage and
abstracting over the clients and servers differences without trading this ease of engineering for
performance? This article presents high-level abstractions defined as deep embedded DSLs in Scala,
that can (1) generate efficient code leveraging the target platform characteristics, (2) be shared
between clients and servers. Though code written with our DSL has a high level of abstraction, our
benchmark on a real world application reports that it runs as fast as hand tuned low-level
JavaScript code.
\end{abstract}

@@

Kossakowski \etal introduced \emph{js-scala}, a compiled embedded DSL defined in Scala that
generates JavaScript code, making it possible to write the client-side code of Web applications
using Scala~\cite{Kossakowski12_JsDESL}. However, Kossakowski \etal did not address the
engineering dilemma described above. This paper enriches js-scala with the following
contributions\footnote{The code is available at
\href{http://github.com/js-scala}{http://github.com/js-scala}}:

@@

We validate our approach with a case study implemented with various candidate technologies and
discuss the relative pro and cons of them. Though the code written in our DSL is high-level and can
be shared between clients and servers, it has the same runtime performance on client-side as
hand-tuned low-level JavaScript code.

@@

Another approach consists in defining several independent domain-specific
languages~\cite{Van00_DSL}, each one focusing on concerns specific to a given problem domain, and
then to combine all the source artifacts written with these language into one executable program, as
shown in figure~\ref{fig:langs} (b). Defining such languages requires a minimal effort compared to
the previous approach because each language has a limited set of features. On the other hand, it is
difficult to have interoperability between DSLs. \cite{Visser07} gave an example of such
a domain-specific language for defining Web applications.

@@

A high-level abstraction for searching elements in a document could be just one function
finding all elements matching a given CSS selector. In fact, most JavaScript
developers\footnote{According to
\href{http://trends.builtwith.com/javascript}{http://trends.builtwith.com/javascript}, jQuery is
used by more than 40\% of the top million sites.} use the jQuery library~\cite{Bibeault08_jQuery}
that actually provides only one function to search for elements. Listing \ref{jquery-selectors}
shows an equivalent JavaScript program as listing \ref{vanilla-selectors}, but using jQuery. The
code is both shorter and simpler, thanks to its higher level of abstraction. jQuery provides an API
that is simpler to master because it has fewer functions, but this benefit comes at the price of a
decrease in runtime performance.

Instead, we propose a solution that has a high-level API but generates JavaScript code using the
specialized native API, when possible, in order to get both ease of engineering and performance.
We achieve this by analyzing, during the first evaluation step, the selector that is
passed as parameter and, when appropriate, by producing JavaScript code using the specialized API,
and otherwise producing code using \code{querySelector} and \code{querySelectorAll}.

@@

Our goal is to evaluate the level of abstraction provided by our solution and its performance, by
comparing it with common approaches. We take the number of lines of code as an inverse approximation
of the level of abstraction. We also evaluate the ability to share code between client and server
sides.

@@

The runtime performance of the vanilla JavaScript, HaXe and js-scala versions are similar (though
the js-scala version is slightly slower by 6\%). It is worth noting that the vanilla JavaScript and
the HaXe versions use low-level code compared to js-scala, as shown in the middle of the figure
(lines of code): the js-scala version needs only 74 lines of code while the vanilla JavaScript
version needs 116 lines of code (57\% bigger) and the HaXe version needs 148 lines of code (100\%
bigger). The jQuery JavaScript version, which code is high-level (54 lines of code, 27\% less
than js-scala) runs 10 times slower than the js-scala version.

@@

We were able to share some DOM fragment definitions between server-side and client-side only in
the js-scala version. In the GWT version we don’t have a choice: dynamic DOM fragments are always
built only on client-side (a practice that makes it more difficult to make the pages content
crawlable by search engines and may increase the initial display time). In the other versions the
code for building the DOM fragment is duplicated between client and server sides, representing 20
lines of JavaScript code (17\% of the total) and 15 lines of HTML (5\% of the total) in the
JavaScript version, and 19 lines of HaXe code (13\% of the total) and 15 lines of HTML (5\%
of the total) in the HaXe version. In the js-scala version the DOM fragment definitions shared
between clients and servers represent 22 lines of Scala code (30\% of the total) and save 15 lines
of HTML (5\% of the total).

@@

Our goal was to put the runtime performance in perspective with the level of abstraction. We are
aware that the indicator we chose as an inverse approximation of the abstraction level, the number
of lines of code, is not scientifically established and may be subject to discussion. However, we
think it is a reasonable approximation in our case because all the candidate languages we use have
a similar syntax, inherited from the C programming language.

@@

We implemented our solution as compiled embedded DSLs in Scala. Generating code from our DSLs is a
two step process: an initial Scala program first evaluates to an intermediate representation of
the final program that is traversed by code generators to produce the final JavaScript code.
Domain-specific optimizations can happen during the IR construction (as shown in section
\ref{sec:selectors}) or during the code generation (as shown in section \ref{sec:option}).

@@

These characteristics allowed us to capture some Web programming patterns as high-level
abstractions, making the code of our application simpler to reason about and making some parts of
the code reusable between client and server sides, while keeping execution performance on
client-side as high as if we used hand-tuned low-level JavaScript code.

@@

Our two kinds of benchmarks, (i) micro-benchmarks to evaluate one abstraction and (ii) a benchmark
on a real application using these abstractions, show performance similar to hand-optimized
low-level code.
